---
title: "Accessibility with R5R - Tutorial"
author: "Saraiva, Marcus"
date: "28/02/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

Accessibility metrics measure the ease with which opportunities, such as
jobs, healthcare, and education facilities, can be reached by a traveler from 
a particular location. 

One of the simplest forms of accessibility metrics is the cumulative-opportunities, 
which counts all of the opportunities accessible from each location in less than a
cutoff time. 

This tutorial shows how to calculate accessibility metrics in R using the 
free and open-source [`r5r` package](https://ipeagit.github.io/r5r/index.html).
`r5r`  provides a simple and friendly interface to R5, a very fast and open source Java-based multi-modal routing engine developed by [Conveyal](https://www.conveyal.com/).
R5 stands for [Rapid Realistic Routing on Real-world and Reimagined networks](https://github.com/conveyal/r5).

In this example, we will be using a subset of the Greater London Authority (GLA) 
area comprised by London's Inner Boroughs. 

## Obtaining the input data

Two input datasets are required to build a multi-modal transport network:

1.  road network dataset from OpenStreetMap, used for routing of walking, 
    cycling and driving trips, as well as access to public transport stations 
    and stops;
2.  public transport schedules in GTFS format;

Both datasets are freely available for the Greater London Authority area. The road
network can be downloaded from [Geofabrik](download.geofabrik.de) daily extracts, 
and London's GTFS feeds are provided by the UK's [Department for Transportation](https://data.bus-data.dft.gov.uk/) open data portal. 

This exercise will be restricted to the Inner London area, and the files can be 
downloaded by running the following chunk of code:

```{r}
piggyback::pb_download("inner-london-latest.osm.pbf", 
                       repo = "mvpsaraiva/r5r_gla", 
                       dest = here::here("data_inner"))

gla_gtfs_url <- "https://data.bus-data.dft.gov.uk/timetable/download/gtfs-file/london/"
download.file(gla_gtfs_url, 
              destfile = here::here("data_inner/itm_london_gtfs.zip"))
```
We also need information on the location and number of opportunities in the study
area, as well as the geographic locations of places where people live and that we
want to evaluate. The data used in this tutorial, in geopackage format, is provided
in the `data_inner` folder of this repository. There are two files in that folder:

1.  `od_zones_inner_gla.gpkg`: MSOA zones of inner London boroughs;
2.  `od_points_inner_gla.gpkg`: lat/lon coordinates of the centroids of the MSOAs.

The following sections will show how to load those datasets and calculate the
accessibility at London's MSOAs.

## Setup and Initialisation

### Load necessary libraries

Before we start using `r5r`, we need to increase the memory available to Java. 
This is necessary because, by default, `R` allocates only 512MB of memory for
Java processes, which is not enough for large queries using `r5r`. To
increase available memory to 6GB, for example, we need to set the
`java.parameters` option, as follows:

```{r echo=TRUE, message=FALSE}
options(java.parameters = "-Xmx6G")
```

Then we can load the packages used in this tutorial:

```{r echo=TRUE, message=FALSE, warning=FALSE}
library(r5r)
library(tidyverse)
library(sf)
library(data.table)
library(ggplot2)
library(mapview)
```

### Load input data

The following code loads the  MSOAs geometries and centroids:

```{r include=FALSE}
od_zones <- st_read(here::here("data_inner", "od_zones_inner_gla.gpkg"))
od_points_sf <- st_read(here::here("data_inner", "od_points_inner_gla.gpkg")) %>% 
  st_cast(to = "POINT")
```

The centroids dataset (`od_points_sf`) contains the following required fields:

1. id: unique identifier of each location;
2. jobs: number of jobs in each location (obs: any kind of opportunity can be
   considered, not neccessarily jobs);
3. geom: the POINT geometry of each location, in WGS 84 CRS.

The code below shows the contents of `od_points_sf`:

```{r}
od_points_sf
```

### Visualising input data

We can use the `mapview` package to open an interactive map of the input points:

```{r}
mapview(od_points_sf, zcol="jobs")
```

## Calculating and visualising accessibility

### Build R5 multi-modal network

Finally, we can use the `setup_r5` function from `r5r` to build a routing network
and load it into memory. The `r5r_core` object returned by `setup_r5` contains 
all we need for the next steps.
The code will combine the data on the road network and the schedule from public transport data (GTFS) to produce a multi-modal transport network.

```{r include=FALSE}
r5r_core <- setup_r5(here::here("data_inner"), verbose = FALSE, temp_dir = TRUE)
```

### Calculating accessibility

Now that everything is in place, we can compute accessibility to jobs with a 
single line of R code:

```{r}
access_df <- accessibility(r5r_core,
                           origins = od_points_sf,
                           destinations = od_points_sf,
                           opportunities_colname = "jobs",
                           mode = c("WALK", "TRANSIT"),
                           departure_datetime = as.POSIXct("02-03-2022 09:00:00", 
                                                           format = "%d-%m-%Y %H:%M:%S"),
                           cutoffs = c(30, 45),
                           verbose = FALSE,
                           progress = TRUE)
```

The `accessibility` function above takes the following parameters:

-  `r5r_core`: the variable created by `setup_r5` that contains the R5 routing
   network and engine in memory;
-  `origins` and `destinations`: the points representing origins and destinations 
   of trips. 
-  `opportunities_colname`: the name of the column in the `destination` points
   dataset that contains the opportunities we are calculating access to (in this
   case, `jobs`);
-  `mode`: transport modes to be considered, where `TRANSIT` includes all modes
   of public transportation (such as busses, the underground, commuter trains,
   and so on);
-  `departure_datetime`: the day and time of the start of the trips. For public
   transport trips, the date must be inside the range available in the GTFS
   data provided. 
-  `cutoffs`: the travel times, in minutes, that we are calculating cumulative-
   opportunities accessibility metrics.
-  `verbose = FALSE`: reduces the amount of outputs to the R console;
-  `progress = TRUE`: shows the progress of the calculation;
   
### Visualising results

Finally, to see the results, we need to join the results of the `accessibility` 
function with the geometry of the input zones. This can be done in R as well:

```{r}
access_sf <- left_join(od_zones, access_df, by = c("code"="from_id"))
```

We can plot the results and visualise how many jobs people can access in 30 and
45 minutes of travel time, using London's public transport system.

```{r}
access_sf %>%
  ggplot(aes(fill=accessibility/1000)) +
  geom_sf(aes(geometry=geom), size=0.2) +
  coord_sf(datum=NA) +
  scale_fill_distiller(palette = "Spectral") +
  labs(fill = "Accessibility\nJobs (x1000)") +
  theme_light() +
  theme(legend.position = "bottom") +
  facet_wrap(~cutoff, labeller = labeller(cutoff = function(s) return(paste(s, "minutes"))))
```
We will now save the maps produced as png images in your folder:

```{r}
ggsave(here::here("output", "map_inner_gla_access.png"), dpi = 300, width = 20, height = 15, units = "cm")
```

and save the shapefile containing the accessibility results to your folder, so you can open it using QGIS and create alternative maps.
Try to open the file and produce a choropleth (graduated colours) map using 10 classes and 'natural breaks' classification method.

```{r}
st_write(access_sf, here::here("output", "inner_gla_access.shp"))
```

Now that you have completed this tutorial, you can try changing the parameters of the accessibility and checking the changes in output. 
Here are some of the changes you can try using the dataset for inner_London:

- Cut-off times: We have calculated the accessibility for 30 and 45 min. Why not try different times? 60 and 90 min cutoff time are commonly used in accessibility analysis. But you can also try 15 min to see how it changes?
- transport modes: The above calculated used 'transit', which means a combination of all available public transport and walking. You can change transit to bus, for example, to see the impact. Or calculate accessibility by using car, walk or bicycle modes
- departure time: you can also change the time and see if it affects your results

